Co-methylation review

### 2014-12-09 Rmarkdown/comethylation_review.Rmd
Minor fixes
Began revising co-methylation review

---
title: "Co-methylation review"
author: "Peter Hickey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

@@

Co-methylation is the dependence structure of DNA methylation data. This dependence structure can be viewed from different perspectives and on different scales, which has led to several disparate definitions of co-methylation. In this chapter I first review these definitions and put them in the statistical framework of Chapter \ref{chap:stat_model}. I highlight various shortcomings of proposed methods, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}. I also review how methods to detect differential methylation have sought to account for and exploit co-methylation.

@@

I define co-methylation as the dependence structure of DNA methylation data. This broad definition encompasses several concepts that have variously been described as "co-methylation", or similar, some of which are closely related and others that are disparate. My definition is broader than "the presence of methylation over a stretch of neighbouring CpG positions" (\cite{Schatz:2004cha}) or "the relationship between the degree of methylation over distance" (\cite{Eckhardt:2006gh}); it also includes the "correlation of two [loci] across many sample" (\cite{Akulenko:2013fl}) and the "_vertical_ (i.e., progenitor to descendant) correlation between the same CpG site in neighboring cell types" (\cite{Capra:2014uh}). A first important point to emphasise is that there exists both within-sample co-methylation and between-sample co-methylation.

Within-sample co-methylation is the spatial dependence of DNA methylation along the genome within a single sample. Many researchers have observed that there is a strong spatial dependence of DNA methylation along a genome, that is, methylation loci near to one another in the genome tend to be similarly methylated (e.g. \cite{Eckhardt:2006gh, Cokus:2008fc, Lister:2009hy, Lacey:2009by, Li:2010fb, Lyko:2010drm, Landan:2012kp}). Moreover, these studies found that the strength of this dependence itself depends on the distance between the loci and on the genomic context. The fundamental level of within-sample co-methylation is the dependence structure of DNA methylation events occurring on the same DNA fragment, which I call _within-fragment co-methylation__. At a higher level is the _correlation of aggregate methylation levels_, such as the autocorrelation of $\beta$-values within a sample.

Between-sample co-methylation is the relationship between methylation levels at a pair of loci across a set of samples[^concordance]. For example, \cite{Akulenko:2013fl} reported 187 pairs of genes whose methylation level was highly correlated (Pearson $|r| \geq 0.75$) across more than 300 breast cancer samples. Also note, however, that measures of within-sample co-methylation can themselves be compared between samples, which is discussed at the end of this chapter.

[^concordance]: One "between-sample correlation of methylation levels" measure that does not fall under my definition of co-methylation is the following. Consider a pair of samples, $j$ and $j'$, and compute $cor(\{\beta_{i, j}\}_{i = 1}^{i = N_{loci}}, \{\beta_{i, j'}\}_{i = 1}^{i = N_{loci}})$. This measure often reported as evidence for the "concordance" or "replicability" of methylation levels between replicates (e.g., when comparing the genome-wide methylation levels of a cell line passaged 4 versus 5 times \cite{Lister:2009hy} or when comparing different technologies and assays \cite{Zhang:2013uu, Stevens:2013hv, Hansen:2011gu}) but is not the same as co-methylation.

There have been several attempts to analyse co-methylation, most with the aim of better understanding the biological processes that lay down and regulate DNA methylation. An increased understanding of co-methylation could also lead to cheaper and more efficient assays of DNA methylation. For example, we could identify methylation loci whose methylation status is highly predictive of the methylation level of a surrounding region and design an assay to interrogate those highly predictive loci. This idea is analogous to the use of tag-SNPs on whole-genome genotyping microarrays.

There are also reasons for studying co-methylation that are less to do with improving our biological understanding and more to do with improving statistical techniques for analysing DNA methylation data. For example, if co-methylation can be better estimated and understood, then statistical techniques can be developed that explicitly account for and even exploit these dependencies to create more powerful and efficient analysis tools.

@@

To date, most analyses of co-methylation have been based on analysing aggregate measurements of methylation, such as $\beta$-values, within individual samples \cite{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}. These analyses of within-sample correlations of aggregate methylation levels have an implicit assumption of stationarity along the genome (__TODO: Discuss with Terry__).

The analysis in \cite{Lister:2009hy} takes a different approach. While it also uses within-sample aggregate methylation levels, rather than estimating correlations of the methylation levels, the autnors instead looks for patterns in the spacing of "methylcytosines" along the genome.

In reviewing this previous work, I have been forced to make my "best guess" about some of the methods since they are often not sufficiently described in the original publication.

\subsection{\cite{Eckhardt:2006gh}}\label{sec:eckhardt}

\cite{Eckhardt:2006gh} is one of the first publications to apply bisulfite-sequencing to a large number of regions and samples. Of particular relevance to a discussion of co-methylation, \cite{Eckhardt:2006gh} has been cited as evidence that "methylation levels are strongly spatially correlated" \cite{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \cite{Hansen:2011gu}. \cite{Eckhardt:2006gh} performed  PCR-amplicon bisulfite-sequencing of 2,524 amplicons on human chromosomes 6, 20 and 22.  PCR-amplicon bisulfite-sequencing is a labour-intensive assay and so it was a considerable effort to produce these data. Nonetheless, it is important to bear in mind that these data are nowhere near as comprehensive as data generated using modern-day whole-genome bisulfite-sequencing assays. Samples were provided by male and female donors across a wide variety of tissue types and ages, and were were generated to begin a catalogue of CpG methylation in various tissue types.

As part of their analysis, the authors analyse co-methylation, which they define as "the relationship between the degree of methylation over distance". They state that "we were able to establish a significant correlation for comethylation over short distances ($\leq 1,000$ bp), it deteriorated rapidly for distances $ >2,000$ bp" and reference Figure 3a in support of this claim. However, what is plotted in Figure 3a is not "correlation" in the usual sense of the word.

In Figure 3a, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp (__TODO: Check permission on re-printing this figure__). Nowhere in the original paper is it clearly described what the "percentage identical methylation" is nor how it is computed[^beck]. The average amplicon length is reported as $411$ bp with a standard deviation of $71$ bp. Therefore, it is impossible for the "percentage identical methylation" to have been obtained from methylation events co-occuring on the same amplicon, at least for those pairs separated by more than approximately $600$ bp.

@@

My best guess as to the definition of "percentage identical methylation" is that it is the proportion of pairs of identical methylation calls at different CpGs from (generally) different amplicons, that is, $Pr(Z_{h, i} = Z_{h', i'})$. I will refer to my best guess of "percentage identical methylation" as $PIM^{*}$, and now turn to some of its properties.

Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $Z_{i, h}$ is sampled, that is, $Pr(Z_{h, i} = 1) = p_{h}$. For two distinct CpGs sequenced on two (potentially) distinct amplicons, $Z_{h, i}$ and  $Z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(Z_{h, i} = Z_{h', i'})$. Not only may the two CpGs be from distinct amplicons, but in the original analysis the amplicons themselves may be from two distinct samples with different genome-wide average methylation levels. This most general setting is described by $Z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $Z_{h, i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(Z_{h, i}, Z_{h', i'}) = c$. In this setting we can compute the theoretical $PIM^{*}$ as follows:

\begin{align*}
PIM^{*} &= Pr(Z_{h, i} = Z_{h', i'}) \\
        &= Pr(Z_{h, i} = 0, Z_{h', i'} = 0) + Pr(Z_{h, i} = 1, Z_{h', i'} = 1) \\
        &= Pr(Z_{h', i'} = 0 | Z_{h, i} = 0) Pr(Z_{h, i} = 0) +
           Pr(Z_{h', i'} = 1 | Z_{h, i} = 1) Pr(Z_{h, i} = 1) \\
        &= \big [ (1 - p_{h'}) + c (1 - p_{h}) \big ] (1 - p_{h}) +
           \big [ (p_{h'} + c (1 - p_{h})) \big ] p_{h} \\
        &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} + 2 c p_{h} (1 - p_{h})
\end{align*}

This result is proved in __TODO: Cite and link to appendix__.

@@

In general, $PIM^{*} \in [0, 1]$. However, there are a few important special cases that are worth discussing:

1. $PIM^{*}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} \in [0, 1]$: Independent and non-identically distributed (_inid_), i.e., $Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
2. $PIM^{*}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c \neq 0$.
3. $PIM^{*}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.

As for $PIM^{*}$, $PIM^{*}_{inid} \in [0, 1]$. In contrast, it is easy to show by differentiation that $PIM^{*}_{iid} \geq 0.5$ and that $PIM^{*}_{did} \geq 0.5 (c + 1)$, which is $> 0.5$ (resp. $< 0.5$) for $c > 0$ (resp. $c < 0$). __TODO: Illustrate these special cases with a plot.__

To return to the plot shown in Figure 3a; the "percentage identical methylation" asymptotes at $40 \%$, or $0.4$, which is less than the theoretical lower-bound of $PIM^{*}_{iid}$. If we believed that all samples had the same average CpG methylation levels, $p_{h}$, then we would conclude that the methylation states of two CpGs separated by a large distance are in fact __anti-correlated__ ($c < 0$) for CpGs separated by approximately $1,000$ bp. However, since we know that there is sample-to-sample variability in the average level of CpG methylation, the more likely explanation for this result is that the $p_{h}$ vary between samples and therefore $PIM^{*} < 0.5$ are possible, even supposing $c > 0$.

Of course, the above discussion pre-supposes that what is labelled "percentage identical methylation" in Figure 3a is mathematically equivalent to $PIM^{*}$. I believe this to be true but due to the aforementioned deficiencies of the method description it is not possible to verify this derivation.

Quite apart from what exactly is plotted in Figure 3a, the analysis is based on data from only $2,524$ amplicons, which is nowhere near a genome-wide analysis. Furthermore, methylation levels have been aggregated across multiple samples and so the profile in Figure 3a does not represent any one sample but a pool of samples.

\subsection{\cite{Cokus:2008fc}}\label{sec:Cokus}

\cite{Cokus:2008fc} published the BS-seq protocol for performing whole-genome bisulfite-sequencing. They demonstrated this technique by generating a genome-wide map of DNA methylation in wildtype  _Arabidopsis Thaliana_. They performed an autocorrelation analysis of the $\beta$-values for CG, CHG and CHH methylation[^arabidopsis]. The autocorrelation computes, for a set of distances ($0-5,000$ nt), the (Pearson) correlation coefficient of methylation levels for all pairs of methylation loci separated by each distance. To emphasise, I believe this analysis uses all pairs of methylation loci, that is, regardless of how many intervening methylation loci there are for each pair.

@@

\cite{Cokus:2008fc} found that methylation levels are highly correlated, particularly for CG methylation levels, and that the correlation decays as a function of genomic distance between methylation loci. From this they concluded that the "significant correlation between methylated cytosines for distances up to 5,000 nucleotides or more [were] probably a reflection of regional foci of methylation throughout the genome and of large blocks of pericentromeric heterochromatin". They also identified correlations between different methylation contexts, e.g. CG vs. CHG, which "suggest[s] complex interactions between the different types of methylation".

The autocorrelation plots showed two periodicities, one of $\approx 167$ nucleotides and the other of $10$ nucleotides (CHH only). Both periodicities are visible in plots of the raw autocorrelations and are confirmed by a Fourier transform analysis of the autocorrelation.

[^CHH_only]: The $10$ nucleotide periodicity was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \cite{Chodavarapu:2010iqa}

@@

The $10$ nucleotide periodicity is equal to the length of one helical DNA turn. From this result Cokus et al. hypothesise that the DOMAINS REARRANGED METHYLASE 2 (_DRM2_), "the main enzyme controlling asymmetric methylation in _Arabidopsis_", contains two active sites that methylate sites 8-10 apart[^comethylation_def2], like its mammalian homologue, DNA methyltransferase 3 (_DNMT3_) (__TODO: If I want a picture explaining this then Figure 3b from \cite{Jurkowska:2011dt} might form the basis for my own illustration). However, in a later paper by the same group, this $10$ nucleotide periodicity was also observed for CG and CHG methylation, at least for nucleosome bound DNA \cite{Chodavarapu:2010iqa}.

[^comethylation_def2]: This is referred to as "co-methylation", here meaning the methylation of multiple loci in a single event, by \cite{Jurkowska:2011dt}. This article reviews a plausible biochemical model for how _DNMT3a_ could co-methylate two CpGs separated by $8-10$ nucleotides; this model includes both the scenario where both CpGs are on the same DNA strand and the scenario where the CpGs are on opposite strands.

The $10$ nucleotide CHH-only periodicity in the autocorrelation analysis of the $\beta$-values is also observed in a _within-fragment analysis_ (discussed in \ref{sec:previous_within-fragment_analyses})

\subsection{\cite{Chodavarapu:2010iqa}}{sec:chodavarapu}

\cite{Chodavarapu:2010iqa} use MNase-seq to study nucleosome positioning and the relationship to DNA methylation, including co-methylation, in _Arabidopsis_ and an embryonic human stem cell line (HSF1). The authors plot a "weighted average methylation", which I presume to be a weighted version of the $\beta$-values, against the distance from predicted nucleosome start sites, which are obtained via MNase-seq. The nucleosome start sites anchor the methylation observations to a common grid and is used to show that nucleosome bound DNA has a greater level of methylation than DNA that is not bound to nucleosomes. A $10$ nucleotide periodicity is readily observed in these plots, consistent with the discovery of $10$ nucleotide periodicities in the autocorrelation of $\beta$-values published by \cite{Cokus:2008fc}. Again, the observed periodicity is confirmed by analysing a Fourier transform of these plots.

In addition to the genome-wide analyses in both _Arabidopsis_ and human embryonic stem cell line (HSF1), the same method is used to examine promoters, genes, repeats, euchromatic regions and centromeric regions of _Arabidopsis_ and promoters, genes, repeats and CpG islands of the HSF1 sample, which confirmed the $10$ nucleotide periodicity. This led \cite{Chodavarapu:2010iqa} to propose an extended version of their earlier hypothesis \cite{Cokus:2008fc}: namely, "that nucleosomes are to some extent dictating access to the DNA and therefore setting the register of methylation __for all DNA methyltransferases__" (emphasis mine) and not just for the DRM2 methyltransferase.

\subsection{\cite{Li:2010fb}}\label{sec:li}

\cite{Li:2010fb} compute "the correlation of methylation level of any two nearby CpGs and the relationship between spatial distance (from one CpG to another) and strength of this correlation". The authors conclude that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $> 1,000$ bp", citing their results as being consistent with those in \cite{Eckhardt:2006gh}. In addition to this genome-wide analysis, \cite{Li:2010fb} perform the same analysis for CpGs within 19 different genomic contexts such as upstream of a gene, untranslated region, coding DNA sequences and various repetitive elements. Somewhat surprisingly, they do not specifically compare CpG islands to non CpG islands, although some of these differences will be captured by the analysis of regions upstream of a gene, which include many CpG island promoters.

Unfortunately, just as in \cite{Eckhardt:2006gh}, \cite{Li:2010fb} never define exactly what is being computed in their analysis of co-methylation and no software is available. All the co-methylation results are presented as supplementary figures, which lack figure legends and are supported by only very brief captions. So, once again, I have been forced to make my best guess as to what these figures show and how it was computed.

Supplementary Figure 6a plots the "correlation" against "distance of CpG" ($0 - 1000$), which I believe are the genome-wide co-methylation estimates over the range $0 - 1000$ bp. Based on the "$n = 18,936,995$" quoted in the header of this figure, which I interpret as the number of pairs used in this plot of "correlation" against "istance of CpG" ($0 - 1000$), I believe that what is shown are the correlations of pairs of CpG $\beta$-values where the pairs are separated by a common distance and have no intervening CpGs. If all pairs were used this $n$ would be much higher since there are more than twenty million CpGs on one strand of the haploid human genome. To emphasise, the correlation shown by \cite{Li:2010fb} is not the same as the autocorrelation computed in \cite{Cokus:2008fc} since the autocorrelation uses all pairs of CpGs, regardless of the number of intervening CpGs.

>>>>

---
author: "Peter Hickey"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

\chapter{A critical review of co-methylation}\label{chap:co-methylation_review}

@@

Co-methylation is the dependence structure of DNA methylation data. This dependence structure can be viewed from different perspectives and on different scales, which has led to several disparate definitions of co-methylation. In this chapter I critically review these definitions and put them in the statistical framework of Chapter \ref{chap:wgbs_statistical_framework}. I highlight various shortcomings of proposed methods, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}. I also review how methods to detect differential methylation have sought to account for and exploit co-methylation.

@@

I define co-methylation as the dependence structure of DNA methylation data. This broad definition encompasses several concepts that have variously been described as 'co-methylation', or similar. My definition is broader than "the presence of methylation over a stretch of neighbouring CpG positions" \citep{Schatz:2004cha}) or "the relationship between the degree of methylation over distance" \citep{Eckhardt:2006gh}; it also includes the "correlation of two [loci] across many sample" \citep{Akulenko:2013fl} and the "_vertical_ (i.e., progenitor to descendant) correlation between the same CpG site in neighbouring cell types" \citep{Capra:2014uh}. A first important point to emphasise is that there exists both within-sample co-methylation and between-sample co-methylation.

Within-sample co-methylation is the spatial dependence of DNA methylation along the genome within a single sample. Many researchers have observed that there is a strong spatial dependence of DNA methylation along a genome, that is, methylation loci near to one another in the genome tend to be similarly methylated \citep[e.g.,]{Eckhardt:2006gh, Cokus:2008fc, Lister:2009hy, Lacey:2009by, Li:2010fb, Lyko:2010dr, Landan:2012kp}. Moreover, these studies found that the strength of this dependence itself depends on the distance between the loci and on the genomic context.

The fundamental level of within-sample co-methylation is the dependence structure of DNA methylation events occurring on the same DNA fragment, which I call _within-fragment co-methylation__. At a higher level is the _correlation of aggregate methylation levels_, such as the spatial correlation of $\beta$-values within a sample.

Between-sample co-methylation is the relationship between methylation levels at a pair of loci across a set of samples[^concordance]. For example, \citet{Akulenko:2013fl} reported 187 pairs of genes whose methylation level was highly correlated (Pearson $|r| \geq 0.75$) across more than 300 breast cancer samples. Also note, however, that measures of within-sample co-methylation can themselves be compared between samples, which is discussed at the end of this chapter.

[^concordance]: One measure of between-sample correlation of methylation levels that does not fall under my definition of co-methylation is the following: Consider a pair of samples, $j$ and $j'$, and compute the correlation between the set of $\beta$-values for sample $j$ against sample $j'$, i.e., $cor(\{\beta_{i, j}\}_{i = 1}^{i = N_{loci}}, \{\beta_{i, j'}\}_{i = 1}^{i = N_{loci}})$. This measure is often reported as evidence for the "concordance" of methylation levels between replicates, but is not what I consider a form of co-methylation.

There have been several attempts to analyse co-methylation, most with the aim of better understanding the biological processes that lay down and regulate DNA methylation. An increased understanding of co-methylation could also lead to cheaper and more efficient assays of DNA methylation. For example, we might identify methylation loci whose methylation status is highly predictive of the methylation level of a surrounding region and design an assay to interrogate those highly predictive loci. This idea is analogous to the use of tag-SNPs on whole-genome genotyping microarrays.

There are also reasons for studying co-methylation that are less to do with improving our biological understanding and more to do with improving statistical techniques for analysing DNA methylation data. If co-methylation can be better estimated and understood, then statistical techniques can be developed that explicitly account for and even exploit these dependencies to create more powerful and efficient analysis methods.

@@

To date, most analyses of co-methylation have been based on analysing correlations of aggregate levels of methylation, such as $\beta$-values, within individual samples \citep{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}. An exception is the analysis in \citet{Lister:2009hy}. While it also uses within-sample aggregate methylation levels, rather than estimating correlations of the methylation levels, the authors instead look for patterns in the spacing of 'methylcytosines' along the genome[^Lister_methylcytosines].

[^Lister_methylcytosines]: See Section \ref{sec:beta-value_interpretation} for a critique of this definition of 'methylcytosine'.

In reviewing this previous work, I have been forced to make my best guess about some of the methods since they are often not sufficiently described in the original publication.

\subsection{\citet{Eckhardt:2006gh}}\label{sec:eckhardt}

\citet{Eckhardt:2006gh} is one of the first publications to apply bisulfite-sequencing to a large number of regions and samples. Of particular relevance to a discussion of co-methylation, \citet{Eckhardt:2006gh} has been cited as evidence that "methylation levels are strongly spatially correlated" \citep{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \citep{Hansen:2011gu}. \citet{Eckhardt:2006gh} performed PCR-amplicon bisulfite-sequencing of $2,524$ amplicons on human chromosomes $6$, $20$ and $22$. PCR-amplicon bisulfite-sequencing is a labour-intensive assay and so it was a considerable effort to produce these data. Nonetheless, it is important to bear in mind that these data are nowhere near as comprehensive as data generated using modern-day whole-genome bisulfite-sequencing assays. Samples were provided by male and female donors across a wide variety of tissue types and ages, and were were generated to begin a catalogue of CpG methylation in various tissue types.

As part of their analysis, the authors analyse co-methylation, which they define as "the relationship between the degree of methylation over distance". They state that "we were able to establish a significant correlation for comethylation over short distances ($\leq 1,000$ bp), it deteriorated rapidly for distances $>2,000$ bp" and reference the plot reprinted in Figure \ref{fig:Eckhardt} in support of this claim. In Figure \ref{fig:Eckhardt}, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp, which is not a correlation in the usual sense of the word. Nowhere in the original paper is it clearly described what the "percentage identical methylation" is nor how it is computed[^beck].

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{../figures/ng1909-F3a.jpg}
\caption{Correlation of DNA methylation with spatial distance. (a) Correlation between comethylation and spatial distance. Orange dots represent CpG methylation values aggregated and averaged over $25,000$ individual measurements. Gray dots represent CpG methylation values based on resampling of random CpG positions. Blue dots indicate CpG methylation values based on resampling of amplicon positions. At distances $>1,000$ bp, we did not detect any correlation between CpG methylation and spatial distance. Adapted by permission from from Macmillan Publishers Ltd: [Nature Genetics] (\citet{Eckhardt:2006gh}), copyright (2006)}
\label{fig:Eckhardt}
\end{figure}

@@

The average amplicon length is reported as $411$ bp with a standard deviation of $71$ bp. Therefore, it is impossible for the "percentage identical methylation" to have been obtained from methylation events co-occuring on the same amplicon, at least for those pairs separated by more than approximately $600$ bp.

My best guess as to the definition of "percentage identical methylation" is that it is the proportion of pairs of identical methylation calls at different CpGs from (generally) different amplicons, that is, $Pr(Z_{h, i} = Z_{h', i'})$. I will refer to my best guess of "percentage identical methylation" as $PIM^{\ast}$, and now turn to some of its properties.

Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $Z_{i, h}$ is sampled, that is, $Pr(Z_{h, i} = 1) = p_{h}$. For two distinct CpGs sequenced on two (potentially) distinct amplicons, $Z_{h, i}$ and  $Z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(Z_{h, i} = Z_{h', i'})$. Not only may the two CpGs be from distinct amplicons, but in the original analysis the amplicons themselves may be from distinct samples with different genome-wide average methylation levels. This most general setting is described by $Z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $Z_{h, i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(Z_{h, i}, Z_{h', i'}) = c$. In this setting we can compute the theoretical $PIM^{\ast}$ as follows:

\begin{align*}
PIM^{\ast} &= Pr(Z_{h, i} = Z_{h', i'}) \\
        &= Pr(Z_{h, i} = 0, Z_{h', i'} = 0) + Pr(Z_{h, i} = 1, Z_{h', i'} = 1) \\
        &= Pr(Z_{h', i'} = 0 | Z_{h, i} = 0) Pr(Z_{h, i} = 0) +
           Pr(Z_{h', i'} = 1 | Z_{h, i} = 1) Pr(Z_{h, i} = 1) \\
        &= \big [ (1 - p_{h'}) + c (1 - p_{h}) \big ] (1 - p_{h}) +
           \big [ (p_{h'} + c (1 - p_{h})) \big ] p_{h} \\
        &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} + 2 c p_{h} (1 - p_{h})
\end{align*}

This result is proven in Appendix \ref{app:bernoulli_equal}.

@@

In general, $PIM^{\ast} \in [0, 1]$ since it is a probability, which is already unlike a correlation coefficient that takes values in $\[-1, 1\]$. There are a few important special cases that are worth discussing:

1. $PIM^{\ast}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'}$: Independent and non-identically distributed (_inid_), i.e., $Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
2. $PIM^{\ast}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c \neq 0$.
3. $PIM^{\ast}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.

As for $PIM^{\ast}$, $PIM^{\ast}_{inid} \in [0, 1]$. In contrast, it is easy to show by differentiation that $PIM^{\ast}_{iid} \geq 0.5$ and that $PIM^{\ast}_{did} \geq 0.5 (c + 1)$, which is $> 0.5$ (resp. $< 0.5$) for $c > 0$ (resp. $c < 0$).

To return to the plot shown in Figure 3a; the "percentage identical methylation" asymptotes at $40 \%$, or $0.4$, which is less than the theoretical lower-bound of $PIM^{\ast}_{iid}$. If we believed that all samples had the same average CpG methylation levels, $p_{h}$, then we would conclude that the methylation states of two CpGs separated by a large distance are in fact __anti-correlated__ ($c < 0$) for CpGs separated by approximately $1,000$ bp. However, since we know that there is sample-to-sample variability in the average level of CpG methylation, the more likely explanation for this result is that the $p_{h}$ vary between samples and therefore $PIM^{\ast} < 0.5$ are possible, even supposing $c > 0$.

Of course, the above discussion pre-supposes that what is labelled "percentage identical methylation" in Figure \ref{fig:Eckhardt} is mathematically equivalent to $PIM^{\ast}$. I believe this to be true but due to the aforementioned inadequacies of the method description it is not possible to verify this derivation.

Quite apart from what exactly is plotted in Figure \ref{fig:Eckhardt}, the analysis is based on data from only $2,524$ amplicons, which is nowhere near a genome-wide analysis. Furthermore, methylation levels have been aggregated across multiple samples and so the profile in Figure \ref{fig:Eckhardt} does not represent any one sample but a pool of samples.

\subsection{\citet{Cokus:2008fc}}\label{sec:Cokus}

\citet{Cokus:2008fc} published the BS-seq protocol for performing whole-genome bisulfite-sequencing. They demonstrated this technique by generating a genome-wide map of DNA methylation in wildtype  _Arabidopsis Thaliana_. They performed an autocorrelation analysis of the $\beta$-values for CG, CHG and CHH methylation[^arabidopsis]. The autocorrelation computes, for a set of distances ($0-5,000$ nt), the (Pearson) correlation coefficient of methylation levels for all pairs of methylation loci separated by each distance. To emphasise, I believe this analysis uses all pairs of methylation loci, that is, regardless of how many intervening methylation loci there are for each pair.

@@

\citet{Cokus:2008fc} found that methylation levels are highly correlated, particularly for CG methylation levels, and that the correlation decays as a function of genomic distance between methylation loci. From this they concluded that the "significant correlation between methylated cytosines for distances up to 5,000 nucleotides or more [were] probably a reflection of regional foci of methylation throughout the genome and of large blocks of pericentromeric heterochromatin". They also identified correlations between different methylation contexts, e.g., CG vs. CHG, which "suggest[s] complex interactions between the different types of methylation".

The autocorrelation plots showed two periodicities, one of $\approx 167$ nucleotides and the other of $10$ nucleotides (CHH only). Both periodicities are visible in plots of the raw autocorrelations and are confirmed by a Fourier analysis of the autocorrelation signal.

[^CHH_only]: The $10$ nucleotide periodicity was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \citep{Chodavarapu:2010iqa}

@@

The $10$ nucleotide periodicity is equal to the length of one helical DNA turn. From this result Cokus et al. hypothesise that the DOMAINS REARRANGED METHYLASE 2 (_DRM2_), "the main enzyme controlling asymmetric methylation in _Arabidopsis_", contains two active sites that methylate sites $8-10$ nucleotides apart[^comethylation_def2], like its mammalian homologue, DNA methyltransferase 3 (_DNMT3_). The $10$ nucleotide CHH-only periodicity in the autocorrelation analysis of the $\beta$-values is also observed in a _within-fragment analysis_ (discussed in \ref{sec:previous_within-fragment_analyses})

[^comethylation_def2]: This is also referred to as "co-methylation" by \citet{Jurkowska:2011dt}, here meaning the methylation of multiple loci in a single event. \citet{Jurkowska:2011dt} reviews a plausible biochemical model for how _DNMT3a_ could co-methylate two CpGs separated by $8-10$ nucleotides; this model includes both the scenario where both CpGs are on the same DNA strand and the scenario where the CpGs are on opposite strands.

\subsection{\citet{Chodavarapu:2010iqa}}{sec:chodavarapu}

\citet{Chodavarapu:2010iqa} use MNase-seq to study nucleosome positioning and the relationship to DNA methylation, including co-methylation, in _Arabidopsis_ and an embryonic human stem cell line (_HSF1_). The authors plot a "weighted average methylation", which I presume to be a weighted version of the $\beta$-values, against the distance from predicted nucleosome start sites, which are obtained via MNase-seq. The nucleosome start sites anchor the methylation observations to a common grid and this plot is used to show that nucleosome bound DNA has a greater level of methylation than DNA that is not bound to nucleosomes. A $10$ nucleotide periodicity is readily observed in these plots, consistent with the discovery of $10$ nucleotide periodicities in the autocorrelation of $\beta$-values published by \cite{Cokus:2008fc}. Again, the observed periodicity is confirmed by analysing a Fourier transform of this correlation signal.

In addition to the genome-wide analyses in both _Arabidopsis_ and human embryonic stem cell line (HSF1), the same method is used to examine promoters, genes, repeats, euchromatic regions and centromeric regions of _Arabidopsis_ and promoters, genes, repeats and CpG islands of the _HSF1_ sample, which confirmed the $10$ nucleotide periodicity. In fact, these results revealed that the $10$ nucleotide is common to all methylation types, including symmetric CpG methylation, at least for nucleosome-bound DNA. This led \citet{Chodavarapu:2010iqa} to discard their earlier hypothesis as to the source of the $10$ nucleotide periodicity \citep{Cokus:2008fc} in favour of one that "nucleosomes are to some extent dictating access to the DNA and therefore setting the register of methylation __for all DNA methyltransferases__" (emphasis mine) and not just for the DRM2 methyltransferase.

\subsection{\citet{Li:2010fb}}\label{sec:li}

\citet{Li:2010fb} compute "the correlation of methylation level of any two nearby CpGs and the relationship between spatial distance (from one CpG to another) and strength of this correlation". The authors conclude that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $>1,000$ bp", citing their results as being consistent with those in \citet{Eckhardt:2006gh}. In addition to this genome-wide analysis, \citet{Li:2010fb} perform the same analysis for CpGs within 19 different genomic contexts such as upstream of a gene, untranslated region, coding DNA sequences and various repetitive elements. Somewhat surprisingly, they do not specifically compare CpG islands to non CpG islands, although some of these differences will be captured by the analysis of regions upstream of a gene, which include many CpG island promoters.

Unfortunately, just as in \citet{Eckhardt:2006gh}, \citet{Li:2010fb} never define exactly what is being computed in their analysis of co-methylation and no software is available. All the co-methylation results are presented as supplementary figures, which lack figure legends and are supported by only very brief captions. So, once again, I have been forced to make my best guess as to what these figures show and how it was computed.

Figure \ref{fig:Li6} plots the "correlation" against "distance of CpG" ($0 - 1000$), which I believe are the genome-wide co-methylation estimates over the range $0 - 1000$ bp. Based on the "$n = 18,936,995$" quoted in the header of panel (a) of this figure, which I interpret as the number of pairs used in this plot of "correlation" against "distance of CpG" ($0 - 1000$), I believe that what is shown are the correlations of pairs of CpG $\beta$-values where the pairs are separated by a common distance and have no intervening CpGs. If all pairs were used this $n$ would be much larger since there are more than twenty million CpGs on one strand of the haploid human genome. To emphasise, the correlation shown by \citet{Li:2010fb} is not the same as the autocorrelation computed in \citet{Cokus:2008fc} since \citet{Li:2010fb} only use pairs of adjacent CpGs whereas \citet{Cokus:2008fc} use all pairs of CpGs, regardless of the number of intervening CpGs.
